---
title: "COVID-19 Deaths by Race/Ethnicity"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, message = F, warning = F}
library(tidyverse)
library(readr)
library(rvest)
library(httr)
library(lubridate)
library(plotly)

knitr::opts_chunk$set(
  fig.width = 10,
  fig.asp = .8,
  out.width = "90%", 
  echo = FALSE,
  warning = FALSE
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```


January 1, 2020 to November 13, 2021

```{r, message = F, warning = F}
death_race_age <- GET("https://data.cdc.gov/resource/ks3g-spdg.csv") %>% 
  content("parsed") %>%
  mutate(start_week = ymd(start_week),
         end_week = ymd(end_week))



#combined set
death_race_age_df = 
  death_race_age %>%
  select(state, race_and_hispanic_origin, age_group_new, covid_19_deaths) %>% 
  filter(race_and_hispanic_origin != "Total Deaths",
         state == "United States",
         age_group_new == "18-29 years" |
           age_group_new == "30-49 years" |
           age_group_new == "50-64 years" | 
           age_group_new == "65-74 years" | 
           age_group_new == "75-84 years" | 
           age_group_new == "85 years and over") %>% 
  group_by(race_and_hispanic_origin, age_group_new) %>% 
  summarize(deaths = sum(covid_19_deaths, na.rm = TRUE))
```

## Graph of deaths by race 

```{r}
covideath <- 
death_race_age_df %>% 
  mutate(age_group = fct_reorder(age_group_new, age_group_new)) %>% 
  rename(`Age Group` = age_group, 
         Deaths = deaths, 
         `Race/origin` = race_and_hispanic_origin) %>% 
  mutate(`Race/origin` = as.factor(`Race/origin`), 
         `Race/origin` = fct_relevel( `Race/origin`, 
                                     levels= c("Unknown", "Non-Hispanic Hative Hawaiian or Other Pacific Islander", "Non-Hispanic More than one race", "Non-Hispanic American Indian or Alaska Native","Non-Hispanic Asian","Non-Hispanic Black","Hispanic", "Non-Hispanic White"))) %>% 
ggplot(aes(x = `Race/origin`, y = Deaths, fill = `Age Group`)) + 
  geom_col(position = "dodge") + 
  labs( title = "COVID-19 Deaths by Race and Hispanic Origin", subtitle = "January 2020 - Novemember 13, 2021") +
  theme(plot.title = element_text(hjust= 1), 
    axis.text.x = element_text(angle = 45), 
    legend.position = "right") + 
  coord_flip() +
  labs(x = "Race and Hispanic Origin",
       y = "COVID-19 Deaths",
       fill = "Age Group")  
```
```{r}
ggplotly(covideath)
```

For Non-Hispanic White and Non-Hispanic Asian, the greatest number of people who died from COVID-19 were in the 85+ year age group. In other minority groups, including Hispanic, Non-Hispanic Black, and Non-Hispanic American Indian or Alaska Native, younger age groups had the highest number of COVID-19 deaths, 50-64 years, 65-74 years, and 50-64 years, respectively. These trends could reflect the health disparities that were exacerbated during the pandemic. 

