---
title: "COVID-19 Deaths by Sex"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r setup, message = FALSE}
library(tidyverse)
library(readr)
library(rvest)
library(httr)
library(lubridate)
#us_states_long <- read_rds(us_states_long.rdata)



```

```{r, warning = FALSE, message = FALSE }

death_age_sex <- GET("https://data.cdc.gov/resource/9bhg-hcku.csv") %>% 
                        content("parsed") %>% mutate(start_date = ymd(start_date), 
                                                      end_date = ymd(end_date))
#combined set
all <- death_age_sex %>% select(start_date, end_date, state,sex,  age_group, covid_19_deaths) %>% 
                         filter(sex == "All Sexes") %>% mutate(start_date = ymd(start_date), 
                                                                end_date = ymd(end_date))



# females only
female <- death_age_sex %>% select(start_date, end_date, state,sex,  age_group, covid_19_deaths) %>% 
                            filter(sex == "Female") %>% mutate(start_date = ymd(start_date), 
                                                                  end_date = ymd(end_date))
# males only 
male <- death_age_sex %>% select(start_date, end_date, state,sex,  age_group, covid_19_deaths) %>% 
                          filter(sex == "Male") %>% mutate(start_date = ymd(start_date), 
                                                            end_date = ymd(end_date))

```

## Graph of deaths by sex 
```{r, message = FALSE, warning = FALSE}
library(viridis)

death_age_sex %>% filter(sex == "Male" | sex == "Female" ) %>% 
                  filter(age_group == "0-17 years"| age_group ==  "18-29 years" |age_group == "30-39 years" 
                         | age_group == "40-49 years" | age_group == "50-64 years"| 
                           age_group ==  "65-74 years"| age_group ==  "75-84 years" | age_group == "85 years and over") %>% 
  mutate(age_group = fct_reorder(age_group, age_group)) %>% 
  ggplot(aes(x = age_group, y = covid_19_deaths, fill= sex)) + 
  geom_col(position = "dodge", stat = "identity") + 
  labs( title = "COVID-19 Deaths by Sex", subtitle = "January 2020 - Novemember 13, 2021") +
  theme(
    axis.text.x = element_text(angle = 90), 
    legend.position = "right", 
    title = element_text(hjust = 0.5)) + 
  xlab("Age Group") + 
  ylab("COVID-19 Deaths") + 
  scale_fill_viridis_d()
  
  
```

For most of the age groups, there have been more COVID-19 deaths in males compared to females. In the oldest age group (85+ years), more women have died from COVID-19 than men. Studies have suggested that there is an association between COVID-19 outcomes and morbidity by sex. These gender differences could potentially be related to the incidence of COVID-19 comorbidities in men and women. Overall, COVID-19 mortality increases with age and there are sex differences in COVID-19 deaths. 
